[Re] Chaos in a long-term experiment with a plankton community

### 2016-02-12 article/article.tex
new article build

The data are available as an Excel file supplement to
\href{http://onlinelibrary.wiley.com/doi/10.1111/j.1461-0248.2009.01391.x/abstract}{an
Ecology Letters publication} \textcite{Beninca2009}. The Excel file
contains several datasheets. Two were particularly important, as they
are the source of the raw data (one contains original species
abundances, the one with the nutrient concentrations). We saved these
two datasheets as comma separated value (csv) text files. In the code
associated with this reproduction, these data files are read from the
associated github repository.

Another datasheet in the Ecology Letters supplement contains transformed
variables (we also saved this as csv file, in order to use it in this
reproduction). We also received a dataset direct from Steve Ellner, see
below for details.

The original species abundance data contained errors (e.g., a few
numerica values had a comma in place of a period as the decimal
separator) that suggested that this was not the exact version of the
dataset used in the original article, or that this was the exact
dataset, but with errors corrected.

@@

The code for this reproduction resides in an R markdown document, as
well as a source file containing some required functions. Some of the
code takes several minutes to run, so an intermediate data file is
provided with results from this code.

@@

The reproduced populations dynamics were at least very similar to those
in figure 1b-g of the original publication (figure \ref{fig:dynamics}).
Note that we plot fourth root transformed values, rather than raw
abundances with a y-axis break, as in the original article.

@@

\subsection{Data transformation}\label{data-transformation}

The following transformation steps were used:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Time series shortened to remove long sequences of zeros.
\item
  Interpolation to create equally spaced observations in time series.
\item
  Fourth root transformation.
\item
  Detrending of five of the time series.
\item
  Rescaling to zero mean and unit standard deviation.
\end{enumerate}

The method for selecting the zeros to remove was unclear. In order to
reproduce this step, we removed the same data as in the original study,
by matching to the transformed data with zeros removed in the Ecology
Letters supplement Excel file mentioned above. All remaining
transformation steps were performed independently of this dataset. Our
reproduction of the transformed data closely matched the published
transformed data.

The data received directly from Stephen Ellner was interpolated, but
without zeros removed. Our interpolated data, without zeros removed,
matched closely this data.

@@

Spectral analyses in the original paper were presented graphically in
figures S3 (raw spectrograms) and S4 (Welch periodograms). These graphs
were, apparently, visually inspected to reveal: ``fluctuations covered a
range of different periodicities'', and ``picophytoplankton, rotifers
and calanoid copepods seemed to fluctuate predominantly with a
periodicity of about 30 days.'' It is unclear how these conclusions were
derived from figures S3 and S4 of the original article. Our reproduced
spectra (not shown here, but code provided) were not quantitatively
identical to the spectra in the original article.

@@

Reproduced divergence rates (figure \ref{fig:divergence}) and Lyapunov
exponents (figure \ref{fig:LE_comparison}) were somewhat different from
those in the original article.. The original article states: ``the
distance between initially nearby trajectories increased over time, and
reached a plateau after about 20--30 days''. The reproduced results
appear not inconsistent with this statement, except for one group of
species (Harpacticoids). The original article also stated that the
analyses ``yielded significantly positive Lyapunov exponents of
strikingly similar value for all species (Fig. 3; mean exponent = 0.057
per day, s.d. = 0.005 per day, n = 9)''. Reproduced exponents had very
similar mean value, but had about four times greater standard deviation
(mean = 0.055 and s.d. = 0.019).

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/LE_comparison.pdf}
\caption{\label{fig:LE_comparison}Comparison of Lyapunov exponents,
estimated by direct method, in the original article and this
reproduction.}
\end{figure}

@@

@@

Although we were not able to make a quantitatively accurate reproduction
of all results of the original article, the qualitative results were
largely identical. For example, all Lyapunov exponents estimated by
direct method are positive, as in the original article, consistent with
chaotic dynamics. Quantitative differences may have resulted from
difference in algorithms used. For example,the original used the
\href{http://www.mpipks-dresden.mpg.de/~tisean/}{Tisean software} to
calculate Lyapunov exponents. As this was available from CRAN
\href{http://cran.r-project.org/web/packages/RTisean/index.html}{until
mid 2014} and since it is a bit less well integrated with R, we instead
use the tseriesChaos package \textcite{tseriesChaos}, which in any case
was largely inspired by the TISEAN project. In addition, there may have
been some difference in algorithm parameters, as not all parameters
required by the functions we used were reported in the original ms.
There may also have been some difference in data used for specific
analyses, e.g., data with zeros removed or not, as it was not always
possible to be totally sure the reproduction used exactly the same data
as the original article.

@@

This reproduction was made as part of the Reproducible Research in
Ecology, Evolution, Behaviour, and Environmental Studies (RREEBES)
Course, lead by Owen Petchey at the University of Zurich. More
information about the course
\href{https://github.com/opetchey/RREEBES/blob/master/README.md}{here}
on github.

>>>>

The data are available as an Excel file supplement to
\href{http://onlinelibrary.wiley.com/doi/10.1111/j.1461-0248.2009.01391.x/abstract}{an
Ecology Letters publication} \textcite{Beninca2009}. The Excel file
contains several datasheets. Two were particularly important, as they
were the source of the raw data (one contains original species
abundances, the other the nutrient concentrations). We saved these two
datasheets, without any alteration, as comma separated value (csv) text
files. In the code associated with this reproduction, these data files
are read from the associated github repository.

Another datasheet (we also saved this as csv file, in order to use it in
this reproduction). We also received a dataset direct from Steve Ellner,
see below for details.

The original species abundance data contained errors (i.e., a few
numeric values had a comma in place of a period as the decimal
separator, and a duplicated date).

The original abundance dataset contains a duplicated date: 28/10/1996
(row 709 and 710 in excel sheet). We changed this (in R) to 26/10/1996,
making the observation occur half way between the two surrounding dates.
The original Protozoa variable in the original dataset contains some
numbers with a comma (instead of period) as the decimal separator; these
were changed to periods.

@@

The code for making the analyses and figures presented in this
reproduction resides in an R markdown document, as well as a source file
containing some required functions. Some of the code takes several
minutes to run, so an intermediate data file is provided with results
from this code.

@@

The reproduced populations dynamics were very similar to those in figure
1b-g of the original publication (figure \ref{fig:dynamics}). As most
analyses were performed on fourth root transformed values, we plot
these, rather than raw abundances with a y-axis break as in the original
article.

@@

\subsection{Data transformations}\label{data-transformations}

The original publication states that long sequences of zeros were
removed from time series prior to transformation and further analysis.
We removed zeros by matching the removals to those in the transformed
version of the data in the published Excel file mentioned above.
Transformation steps were:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Interpolation to create equally spaced observations in time series.
\item
  Fourth root transformation.
\item
  Detrending of five of the time series.
\item
  Rescaling to zero mean and unit standard deviation.
\end{enumerate}

The reproduced transformed data closely matched the published
transformed data (figure \ref{fig:trans_comp1}).

\begin{figure}[htbp]
\centering
\includegraphics{figures/zero_rem_trans_comparison.pdf}
\caption{\label{fig:trans_comp1}Comparison of published transformed data
with reproduced transformed data. The red line is the 1:1 line.}
\end{figure}

@@

Spectral analyses in the original paper were presented graphically in
figures S3 (raw spectrograms) and S4 (Welch periodograms). These graphs
were, apparently, visually inspected to reveal: ``fluctuations covered a
range of different periodicities'', and ``picophytoplankton, rotifers
and calanoid copepods seemed to fluctuate predominantly with a
periodicity of about 30 days.'' It is unclear how these conclusions were
derived from figures S3 and S4 of the original article. Our reproduced
spectra (not shown here, but code provided) were a reasonably close
match to the spectra in the original article.

@@

Reproduced divergence rates (figure \ref{fig:divergence}) and Lyapunov
exponents (figure \ref{fig:LE_comparison}) were somewhat different from
those in the original article.. The original article states: ``the
distance between initially nearby trajectories increased over time, and
reached a plateau after about 20--30 days''. The reproduced results
appear not inconsistent with this statement, except for one group of
species (Harpacticoids). The original article also stated that the
analyses ``yielded significantly positive Lyapunov exponents of
strikingly similar value for all species (Fig. 3; mean exponent = 0.057
per day, s.d. = 0.005 per day, n = 9)''. Reproduced exponents had very
similar mean value, but had much larger standard deviation (mean = 0.057
and s.d. = 0.014).

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/LE_comparison.pdf}
\caption{\label{fig:LE_comparison}Comparison of original and reproduced
directly estimated Lyapunov exponents.}
\end{figure}

@@

The data received directly from Stephen Ellner (file
`interp\_short\_allsystem\_newnames.csv' in the reproduction repository)
was interpolated, but without zeros removed. The reproduced interpolated
data, without zeros removed, matched closely this data (figure
\ref{fig:trans_comp2}).

\begin{figure}[htbp]
\centering
\includegraphics{figures/repro_ellner_comp.pdf}
\caption{\label{fig:trans_comp2}Comparison of interpolated data provided
by Ellner with reproduced interpolated data (no removal of zeros). The
red line is the 1:1 line.}
\end{figure}

@@

The reproduced results were qualitatively identical to those in the
original article. For example, all Lyapunov exponents estimated by
direct method were positive, as in the original article, consistent with
chaotic dynamics. Quantitative differences may have resulted from
difference in algorithms used. For example,the original used the
\href{http://www.mpipks-dresden.mpg.de/~tisean/}{Tisean software} to
calculate Lyapunov exponents. As this was available from CRAN
\href{http://cran.r-project.org/web/packages/RTisean/index.html}{until
mid 2014} and since it is a bit less well integrated with R, we instead
use the tseriesChaos package \textcite{tseriesChaos}, which in any case
was largely inspired by the TISEAN project. In addition, there may have
been some difference in algorithm parameters, as not all parameters
required by the functions we used were reported in the original ms.

Considerable quantitative difference in predictability decay
(\ref{fig:prediction_distance}) remain unexplained.

@@

This reproduction was made as part of the Reproducible Research in
Ecology, Evolution, Behaviour, and Environmental Studies (RREEBES)
Course, lead by Owen Petchey at the University of Zurich. More
information about the course
\href{https://github.com/opetchey/RREEBES/blob/master/README.md}{here}
on github. Many thanks to Elisa Benincà and Steve Ellner for assistance,
and to Steve for providing unpublished data and code.
